GEE models

### 2017-07-25 doc/submission/reviewer-response.md

# Responses to reviewers

## Referee: 1

- *"First, I would like to congratulate you for the exhaustive and complete
research work that has been done for this paper. I think the statistical study
is very successful and useful for researchers who have accumulated a large
database and need a method with the ability to handle a large number of
variables and their interactions. I really liked read in the discussion the
negative points of the study, with which I fully agree, and not an attempt to
hide the limitations of any scientific work. In retrospect, perhaps it will miss
data that clarify the pre-diabetic state, or not, of the patients (HbA1c or
C-peptide). I think it would be good for the study know if some patients were
taking any medication that could affect the results. Although it is a
complicated scientific work to understand for researchers who do not usually use
this kind of statistical studies, I believe that is described in a very
appropriate and clear manner."*

    We thank the reviewer for these comments. In the PROMISE cohort, we 
    collected medication data from all participants. While most participants 
    took some form of medication over the 6 years, many (n=60-103 between the 
    three visits) were on taking lipid and cholesterol controlling medications.
    However, for those who were taking medications that could influence the
    results, additional analyses revealed that there were no differences in the
    predictor or outcome measures with medication use, nor did adjustment for
    medication use substantially contribute to GEE model fit compared to models
    without this variable (see QIC table). We have included this information in
    the revised manuscript.

@@

1. *"Title of the paper: it is a statement of the results. I would prefer to leave
it open ended without an interpretation of the results, as with different levels
of adjustments the results and conclusion will change substantially. See also
comments on the results section."*

    As suggested, we've made the title agnostic.

2. *"Methods in the abstract: a) Please give some information on the cohort
characteristics: age, sex, high risk of diabetes but no prevalent disease at
baseline; b) incident diabetes cases (how were these handled? what about treated
patients? - these should go in the main methods section); c) adjustment for time
– I would prefer to state that time since baseline was used as the underlying
time variable instead of adjusting for time"*

    We have made as many of the requested changes to the methods as the reviewer
    suggested without going over the word limit. Incident diabetes cases are
    presented for descriptive purposes though were not included in any GEE
    analysis. We thank the reviewer for the suggestion regarding time adjustment
    and agree that using time since baseline is a better variable to adjust for
    in the GEE models; the time variable has been replaced in GEE models, though
    the change does not substantially impact the final results. This change has
    been included in the statistical analysis section.

@@

3. *"Results in the abstract: a major problem in my view with the
interpretation of the results is that all characteristics are taken as
confounders although a considerable proportion of time-varying covariates are
taken after the NEFA measurements. Furthermore, it is not clear whether these
measures are confounders, mediators or NEFAs are mediators of these variables. I
think that unadjusted results are as interesting and require discussion as the
fully adjusted and selected models. Regarding the abstract, I would show both
the unadjusted and the fully adjusted results here."*

    The specifics of the covariates have been expanded upon in the methods
    section and the ESM Methods of the revised manuscript. We
    have included the unadjusted results in the main manuscript and have 
    described it in the results and discussion sections, in addition to briefly
    describing them in the abstract (keeping within the word limit). An
    explanation of our interpretation of each covariate in regard to NEFA and the outcomes
    (e.g. mediator or confounder) has been added in the ESM Methods.

@@

4. *"Conclusions in abstract: I have a different interpretation of the results,
particularly because adjustment waist may be an over-adjustment."*

    We have included a sensitivity analysis identifying which exact covariate
    attenuates the results between the unadjusted and adjusted models; the
    covariate is waist circumference. While inclusion of this variable
    attenuates most of the associations, we believe that adjustment of this
    variable is necessary given that biologically NEFA come from the adipose
    tissue during fasting and hypothetically more adipose tissue would indicate
    higher NEFA. Since other aspects of adipose tissue (e.g. role in
    inflammation and appetite regulation) could also impact insulin sensitivity
    and beta-cell function, we believe it is an important confounder to adjust
    for. We have included in the conclusions a discussion on how adjustment of
    waist circumference influences the model and the interpretation, as well as
    show the results of models without waist circumference in the ESM.

@@

5. *"INTRODUCTION: a clear and well-written introduction with a clear objective
and hypothesis"*

6. *"MATERIALS & METHODS:"*
    - *"ESM Figure 1 is not clear to me: a) Prevalent diabetes cases were excluded
    at baseline (n=XX); b) the lines suggest that everyone had to have a year 6
    visit, ehile in the main text, the authors state that any follow-up visit
    was enough for being included (this corresponds well with the GEE
    requirements); c) what happened with incident diabetes cases? Were treated
    diabetes cases included in the analysis? How does that effect measures of
    insulin sensitivity and beta-cell function? I suggest to exclude those
    visits where diabetes was treated."*

        In the revised manuscript, we have clarified how we dealt with incident
        diabetes cases. Briefly, we only report incident diabetes and
        dysglycemia cases in the text to describe the cohort, but in statistical
        analyses and modeling, these cases were excluded. Regarding point a)
        above prevalent diabetes cases at baseline were not included in this
        analysis as not prevalent diabetes cases had fatty acids measured. For
        point b) above, not all participants attended each visit; some
        participants missed the 3-yr visit, but came in for the 6-yr visit. We
        have included wording and numbers within the Results section of the
        revised manuscript regarding the sample and percentage who attended at
        least two visits. Lastly, regarding point c) above because diabetes can
        effect the values of the OGTT-derived indices due to medication the
        patient is taking and due to diabetes itself impacting beta-cell
        function and insulin resistance, we excluded them from all analyses. To
        restate, diabetes cases were not included in GEE or other analyses.

@@

    - *"In the statistical analysis you state that co-variates were
    time-dependent. This is not true for age at baseline, family history, sex,
    ethnicity. Please clarify in the text."* ... *"misspelling: 'statistically
    significance'"*

        Thanks for catching these issues. They have been corrected in the
        revised manuscript.

    - *"I am not an expert of PLS-DA but it is not clear to me, why does this
    analysis reflect more the role of relative NEFA (mole%) analysis than the
    concentration analysis? As I can see, it depends on what variables you feed
    in the model."*

        For the PLS analysis, we used the mol% data to input into the PLS model.
        PLS extracts underlying correlations from high dimensional data
        (multiple predictor variables, i.e. the fatty acids), constrained by a
        outcome variable (i.e. beta-cell function). In this context, because
        mol% data is by definition restricted to sums of 100% (all fatty acid
        values add up to 100%), the underlying correlations between fatty acids
        is dependent on the relative contributions those fatty acids have to the
        total and as such can reflect correlations of specific fatty acids or 
        groups of fatty acids that increase or decrease in proportion together. 
        The concentration data is not contrained by this inherent correlation 
        and thus underlying groups or clusters may not be identified using PLS.

@@

7. *"RESULTS:"*
    - *"Please give ns in Table 1."* ... *"It would be also nice to see, how
    many participants had 1, 2 or 3 time points in the analysis."*

        The number of participants who attended all three visits has been added
        as requested. Regarding including 'ns' in Table 1, we intentionally did
        not include p-values for this table for several reasons. Our objective
        was not to examine significant differences in the basic characteristics
        of the cohort and this table is included to show a purely descriptive
        summary of the cohort over the 6 years. Our analysis currently generates
        a large number of tests and subsequent p-values and, even though we
        adjusted these p-values for the false discovery rate, we did not want to
        generate more p-values (and the resulting increase in Type 1 error) than
        necessary if they are not primary to our objective.

@@

    - *"How were the changes calculated (GEE?) in the basic characteristics of
    the PROMISE cohort section?"*

        The changes over time as reported in the results of the outcome
        variables are simple median differences between the baseline and 6-year
        visit. The p-value representing the difference was computed using an
        unadjusted (time only) GEE model.

    - *"I suggest to include the unadjusted data to the main paper (ESM Figure 3)
    and describe these findings as we do not know the exact model where
    over-adjustment start to happen."*

        The unadjusted results and figure have been added to the main paper and
        described in the results section. We've also added a sensitivity
        analysis confirming which covariate(s) attenuates the associations and
        present the GEE model results of that sensitivity analysis in the ESM.

@@

    - *"In a trajectory analysis, the interaction between time and the main
    predictor (NEFA in this case) is of major interest. The authors check for
    this interaction only as the last step of their model building (see ESM). I
    would suggest to perform this as the first step of the analysis (M0), and
    drop this term, if it is non-significant or does not improve model fit.
    Furthermore, given that this interaction was dropped from the model, it is
    hard to except the major conclusion that NEFA predicts progression of
    beat-cell dysfunction. Actually what is found here is that NEFA are related
    to beta-cell function in repeated cross-sectional analyses and NEFA related
    differences in beta-cell function remain constant over time."*

        The reviewer makes an excellent point regarding the importance of
        conducting an interaction check in the early stages of
        model building. We have incorporated this new approach in the analysis and
        report on the results in the revised paper. In fundamental terms, the reviewer is
        correct that without an interaction term the model is essentially
        "repeated cross-sectional analyses", however, GEE handles the analysis
        with a slightly more nuanced approach. GEE calculates the estimates of the
        association of NEFA with the outcomes at any given time point, taking into
        consideration the inherent multivariate structure of the repeated
        measurements of the outcome variables. As such, inherent similarities in
        the outcome measurements over time within a subject are included in the
        computation of the GEE estimates. This nuance provides substantially
        more power to the model and the proceeding results. See ref 27 in the
        manuscript for more detail on this statistical technique. Within the
        manuscript we have expanded on this nuance to clarify the
        interpretation in the statistical analysis section of the Methods in
        the revised manuscript.

@@

    - *"The model building is based on literature data which is the right way to
    build the models but a more detailed discussion on over-adjustment is
    required before interpretation."*

        To clarify, the model building processes used two additional techniques:
        quasi-likelihood information criteria (QIC) and directed acyclic graphs.
        The results and output of these techniques can be found in the ESM.
        Combining all of these processes in building the model ensures that
        adjustment is empirically based and reduces (but not eliminates) the
        potential for over-adjustment. However, we have added additional
        sensitivity analyses to identify covariates that strongly influence the
        results and discuss the results in light of these findings. We have
        presented the results of excluding the covariate (waist circumferen) in
        the ESM.

@@

    - *"Latent classes. Although this is a very attractive method, based on ESM
    figure 4, there are 3 parallel declining group trajectories that is mostly
    affected by the baseline values. Groups based on baseline (last follow-up)
    tertiles would be easier to interpret."* ... *"The PLS-DA analysis needs
    further description. I would expect variables with the highest loadings to
    be the variables in the unadjusted GEE analysis. I am probably wrong given
    the results."* ... *"Do we really expect good discrimination? The
    associations are week to nul between NEFAs and the outcomes. I do not really
    see the point the extensive supplementary analysis on PLS-DA. Furthermore
    dysglycemia is an outcome that is not directly related to beta-cell function
    only. Dysglycemia requires both insulin resistance and beta-cell
    dysfunction."*

        Given this reviewer's previous comments and suggestions, we have decided
        to: focus on the GEE analysis in the main paper; remove the LCMM
        analysis; perform the PLS analysis on the baseline (cross-sectional)
        data using the outcomes as continuous variables; and, briefly describe
        the findings in the Results section. To keep the ESM material
        streamlined, we have contained the PLS results and other additional
        analyses conducted within the code (accessible via the DOI) for the
        interested reader. We believe this simplifies the results and primary
        findings of this paper without sacrificing a loss in the potential use
        these additional analyses may provide for the interested researcher.

    - *"The authors mention in the Discussion the external validity is mostly
    limited to white females. I suggest to run a sensitivity analysis on this
    group."*

        We agree with the reviewer's comment and is the reason we originally 
        tested for interaction tests by sex and ethnicity. In both cases, we
        found no statistically significant interactions. While these results
        internally suggest no differences between sex or ethnicity, there may
        still be features unique to sex and ethnicity in this cohort that limit
        generalizability. Regardless, in light of the reviewers request, we ran
        stratified models by either sex or ethnicity. In these findings, all 
        significant associations from the non-stratified analyses were 
        attenuated in the stratified analyses, likely due to the reduction in
        sample size in the stratified models. Stratifying analyses has the
        limitation of substantially reducing sample size and is the reason why
        interaction testing is prefered.

@@

        In order to confirm the impact the comment made by the reviewer has on
        the reported associations, we ran sensitivity analyses where we had all
        covariates as time-independent, to match the NEFA variables in the GEE
        model. We found that there were negligible differences in the results
        between models with covariates as either time-dependent and
        time-independent and models with all covariates as time-independent.

# Associate Editor

- *"Many thanks for submitting this potentially interesting paper. The second
reviewer comes with some very helpful comments that will need to be addressed in
detail. There is need to consider the results from the Paris Prospective study:
Charles MA et al.  The role of non-esterified fatty acids in the deterioration
of glucose tolerance in Caucasian subjects: results of the Paris Prospective
Study. Diabetologia 1997;40:1101-6"*

    We thank the associate editor for the kind comments. We have included the
    cited paper in the revised Discussion section.

### 2017-08-08 doc/manuscript.Rmd
updated manuscript word, table, and figure count
added ref to gee esm fig for no waist
minor edits before submission

Word count - Abstract: 242 / 250

Word count - Main text: 3726 / 4000

<!-- Table and figure count: 3 / 4 -->

<!-- Reference limit 39 / 50 -->

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and in addition can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a]. 
Observational and clinical studies have reported concordant findings, showing in
particular that elevated total plasma NEFA associates with an increased risk for
incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].

@@

Sensitivity analyses revealed that adjustment for WC was responsible for 
attenuating the unadjusted results (data not shown). Adjusting for TAG 
attenuated all associations with IGI/IR and ISSI-2 (data not shown), though this
may be due to the fact that only a subset of participants at the 6 year visit 
had TAG measured. Sensitivity analyses to examine the association with the sum 
of fatty acid classes (saturated vs unsaturated) revealed similar findings as 
seen with the adjusted GEE model for total NEFA (data not shown). There were no 
significant interaction effects of time, sex, or ethnicity in either unadjusted
or adjusted models. Medication use for controlling lipid and cholesterol did not
influence the results (data not shown). Raw values from the GEE models are shown
in `r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee_adj')` for
fully adjusted models. Consistent with the GEE modeling, results from modeling
all NEFA in a multivariate environment using PLS showed no discernible
clustering nor was it predictive of the outcomes (data not shown).

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw in general no association of any individual or total NEFA with
hepatic (HOMA2-%S) or whole-body insulin sensitivity (ISI), except for EPA. 
There is some evidence [@Steffen2015a;@Virtanen2014;@Sarbolouki2013] that EPA
may be involved with lowering the risk for diabetes, potentially due to its role
in inflammation [@Calder2009]. However, given we found the association only with
HOMA2-%S and not with ISI, the significant association could be chance finding,
which will require future studies to confirm. We also identified WC as having a
powerful attenuating effect on the associations between unadjusted and adjusted
models. There are some possible explanations for our null findings for the
insulin sensitivity measures and the influence of WC on the results. First, WC
may be a strong causal link in previously reported associations between NEFA and
insulin sensitivity, given the role adipocytes play in metabolism (e.g. leptin,
adiponectin). Second, there may be differences in physiology between fasting and
postprandial NEFA kinetics that we are not able to investigate but that may
explain our null findings for fasting NEFA and insulin sensitivity. For
instance, some experimental studies using clamp protocols found that fasting
NEFA was a weak predictor of insulin sensitivity compared to postprandial
concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
which may be more metabolically active given postprandial activity. Third, the
null findings seen may be due to the high risk population examined in PROMISE, 
as the majority of subjects had a high WC and BMI. It may be that in this
population, insulin resistance has become well established and NEFA may not
contribute to insulin resistance at this somewhat more advanced stage in the
pathogenesis of diabetes.

@@

```{r figure2_gee_adj, fig.cap=fig_gee_adj}
gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    plot_gee_main()
```

>>>>

Word count - Main text: 3869 / 4000 (with figure captions)

<!-- Table and figure count: 4 / 4 -->

<!-- Reference limit 49 / 50 -->

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a]. 
Observational and clinical studies have reported concordant findings, showing in
particular that elevated total plasma NEFA associates with an increased risk for
incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].

@@

Sensitivity analyses revealed that adjustment for WC was responsible for 
attenuating the unadjusted results (see `r cite_sf("gee_adj_nowaist")`). Adjusting for TAG 
attenuated all associations with IGI/IR and ISSI-2 (data not shown), though this
may be due to the fact that only a subset of participants at the 6 year visit 
had TAG measured. Sensitivity analyses to examine the association with the sum 
of fatty acid classes (saturated vs unsaturated) revealed similar findings as 
seen with the adjusted GEE model for total NEFA (data not shown). There were no 
significant interaction effects of time, sex, or ethnicity in either unadjusted
or adjusted models. Medication use for controlling lipid and cholesterol did not
influence the results (data not shown). Raw values from the GEE models are shown
in `r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee_adj')` for
fully adjusted models. Consistent with the GEE modeling, results from modeling
all NEFA in a multivariate environment using PLS showed no discernible
clustering nor was it predictive of the outcomes (data not shown).

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw in general no association of any individual or total NEFA with
hepatic (HOMA2-%S) or whole-body insulin sensitivity (ISI), except for EPA. 
There is some evidence [@Steffen2015a;@Virtanen2014;@Sarbolouki2013] that EPA
may be involved with lowering the risk for diabetes, potentially due to its role
in inflammation [@Calder2009]. However, given we found the association only with
HOMA2-%S and not with ISI, the significant association could be a chance finding,
which will require future studies to confirm. We also identified WC as having a
powerful attenuating effect on the associations between unadjusted and adjusted
models. There are some possible explanations for our null findings for the
insulin sensitivity measures and the influence of WC on the results. First, WC
may be a strong causal link in previously reported associations between NEFA and
insulin sensitivity, given the role adipocytes play in metabolism (e.g. leptin,
adiponectin). Second, there may be differences in physiology between fasting and
postprandial NEFA kinetics that we are not able to investigate but that may
explain our null findings for fasting NEFA and insulin sensitivity. For
instance, some experimental studies using clamp protocols found that fasting
NEFA was a weak predictor of insulin sensitivity compared to postprandial
concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
which may be more metabolically active given postprandial activity. Third, the
null findings seen may be due to the high risk population examined in PROMISE, 
as the majority of subjects had a high WC and BMI. It may be that in this
population, insulin resistance has become well established and NEFA may not
contribute to insulin resistance at this somewhat more advanced stage in the
pathogenesis of diabetes.

@@

```{r figure3_gee_adj, fig.cap=fig_gee_adj}
gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    plot_gee_main()
```
